Iconic representations of molluscs through the ages

### 2012-10-31 large_number_squeezing_with_atoms.tex
Minor appendix fixes

\appendix
\section{Derivation of two mode squeezing formula}
\label{appendixTwoModeDerivation}
In this Appendix we derive the equations for the number variances which were earlier stated without proof. We use the model described in Section \ref{secTwoModeAnalytic}, where we consider a two mode system with $N$ particles, initially prepared as a coherent state with all the population in mode $|a\rangle$ at time $t=0$. A coupling pulse is then applied, so that the system evolves under the Hamiltonian
\begin{equation}
\hat{H_1} = \hbar \omega \hat{a}^{\dagger} \hat{a} +  \hbar \omega \hat{b}^{\dagger} \hat{b} 
          + \Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger} \hat{a} )
\end{equation}
until time $t=t_1$. From time $t_1$ until time $t=t_2$ the coupling is turned off, and the system evolves under the Hamiltonian
\begin{equation}
\hat{H_2} = \frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}.
\end{equation}
Finally, from time $t_2$ until time $t=t_3$, the coupling is once again applied with a phase offset of $\phi$ relative to the first time, and the system evolves via
\begin{equation}
\hat{H_3} = \hbar \omega \hat{a}^{\dagger} \hat{a} +  \hbar \omega \hat{b}^{\dagger} \hat{b}
          + \Omega (e^{i\phi} \hat{a}^{\dagger} \hat{b} + e^{-i\phi} \hat{b}^{\dagger} \hat{a} )
\end{equation}
It is clear that the linear self-energy terms proportional to $\omega$ result in a constant phase rotation of the fields $\hat{a}$ and $\hat{b}$. As this system is self-contained and global phase is unmeasurable, all phase measurements can only rely on the difference in phase between the two modes. This phase rotation is therefore irrelevent and will be ignored in the following analysis.

@@

Beginning with the Hamiltonian $H_1$, during the period $(0, t_1)$ mode $\hat{a}$ will evolve as
\begin{equation}
\hat{a}_1 = e^{ i H_1 t/ \hbar} \, \hat{a}_0 \, e^{-i H_1 t/ \hbar}.
\end{equation}
Utilizing the Hadamard lemma
\begin{equation}
e^X Y e^{-X} = Y + [X,Y] + \frac{1}{2!}[X,[X,Y]] + \frac{1}{3!}[X,[X,[X,Y]]] + \ldots
\label{eqHadamard}
\end{equation}
we obtain 
\begin{eqnarray}
\hat{a}_1 &=& \hat{a}_0 - i \Omega t_1 \hat{b}_0 + \frac{(i \Omega t_1)^2}{2!} \hat{a}_0 - \frac{(i \Omega t_1)^3}{3!} \hat{b}_0 + \frac{(i \Omega t_1)^4}{4!} \hat{a}_2 - \ldots \nonumber \\
          &=& \cos (\Omega t_1) \hat{a}_0 -i \sin (\Omega t_1) \hat{b}_0
\label{eqa1}
\end{eqnarray}
Similarly, we find
\begin{equation}
\hat{b}_1 = \cos (\Omega t_1) \hat{b}_0 - i \sin (\Omega t_1) \hat{a}_0.
\label{eqb1}
\end{equation}
From the form of (\ref{eqa1}) and (\ref{eqb1}) it is clear that as both modes $\hat{a}$ and $\hat{b}$ began in coherent states, they will remain in coherent states, with only their amplitudes changing. Specifically, at time $t_1$ the system is in a product of coherent states, which we denote $|\alpha, \beta\rangle$ with $\alpha=\sqrt{n_a}$ and $\beta=-i\sqrt{n_b}$.

@@

          e^{-i \lambda_{ab} \hat{a}_1^{\dagger} \hat{a}_1 \hat{b}_1^{\dagger} \hat{b}_1}.

@@

   &=& \exp[-i \lambda_{aa} \hat{a}_1^{\dagger} \hat{a}_1] \hat{a}_1

@@

{\mathrm{Var}}[N_b] &=& \langle \hat{b}^{\dagger}_3 \hat{b}_3 \hat{b}^{\dagger}_3 \hat{b}_3 \rangle - \langle \hat{b}^{\dagger}_3 \hat{b}_3 \rangle ^2

@@

\hat{a}^{\dagger}_3 \hat{a}_3 &=& \hat{a}^{\dagger}_2 \hat{a}_2 c^2 +  \hat{b}^{\dagger}_2 \hat{b}_2 s^2 + i c s e^{-i \phi} \hat{b}^{\dagger}_2 \hat{a}_2 - i c s e^{i \phi} \hat{a}^{\dagger}_2 \hat{b}_2 \\

@@

\hat{b}^{\dagger}_3 \hat{b}_3 &=& \hat{b}^{\dagger}_2 \hat{b}_2 c^2 +  \hat{a}^{\dagger}_2 \hat{a}_2 s^2 + i c s e^{i \phi} \hat{a}^{\dagger}_2 \hat{b}_2 - i c s e^{-i \phi} \hat{b}^{\dagger}_2 \hat{a}_2 \\

@@

As $\hat{a}^{\dagger}_1 \hat{a}_1$ and $\hat{b}^{\dagger}_1 \hat{b}_1$ commute with the Hamiltonian $H_2$, they are constants of motion so we have
\begin{equation}
\langle \hat{a}^{\dagger}_2 \hat{a}_2 \rangle = \langle \hat{a}^{\dagger}_1 \hat{a}_1 \rangle = \langle \alpha, \beta | \hat{a}^{\dagger}_1 \hat{a}_1 |\alpha, \beta \rangle = |\alpha|^2 = n_a
\end{equation}
%
\begin{equation}
\langle \hat{b}^{\dagger}_2 \hat{b}_2 \rangle = \langle \hat{b}^{\dagger}_1 \hat{b}_1 \rangle = \langle \alpha, \beta | \hat{b}^{\dagger}_1 \hat{b}_1 |\alpha, \beta \rangle = |\beta|^2 = n_b
\end{equation}
%
\begin{equation}
\langle \hat{a}^{\dagger}_2 \hat{a}_2 \hat{b}^{\dagger}_2 \hat{b}_2 \rangle = \langle \hat{a}^{\dagger}_1 \hat{a}_1 \hat{b}^{\dagger}_1 \hat{b}_1 \rangle = \langle \alpha, \beta | \hat{a}^{\dagger}_1 \hat{a}_1 \hat{b}^{\dagger}_1 \hat{b}_1 |\alpha, \beta \rangle = |\alpha|^2 |\beta|^2 = n_a n_b
\end{equation}
Furthermore, $\hat{a}^{\dagger}_2 \hat{a}_2 \hat{a}^{\dagger}_2 \hat{a}_2$ and $\hat{b}^{\dagger}_2 \hat{b}_2 \hat{b}^{\dagger}_2 \hat{b}_2$ also commute with $H_2$, giving
\begin{equation}
\langle \hat{a}^{\dagger}_2 \hat{a}_2 \hat{a}^{\dagger}_2 \hat{a}_2 \rangle = \langle \hat{a}^{\dagger}_1 \hat{a}_1 \hat{a}^{\dagger}_1 \hat{a}_1 \rangle = \langle \alpha, \beta | \hat{a}^{\dagger}_1 \hat{a}^{\dagger}_1 \hat{a}_1 \hat{a}_1 + \hat{a}^{\dagger}_1 \hat{a}_1 |\alpha, \beta \rangle = n_a^2 + n_a
\end{equation}
%
\begin{equation}
\langle \hat{b}^{\dagger}_2 \hat{b}_2 \hat{b}^{\dagger}_2 \hat{b}_2 \rangle = \langle \hat{b}^{\dagger}_1 \hat{b}_1 \hat{b}^{\dagger}_1 \hat{b}_1 \rangle = \langle \alpha, \beta | \hat{b}^{\dagger}_1 \hat{b}^{\dagger}_1 \hat{b}_1 \hat{b}_1 + \hat{b}^{\dagger}_1 \hat{b}_1 |\alpha, \beta \rangle = n_b^2 + n_b
\end{equation}
The remaining terms are not constants of motion, so we proceed by making use of (\ref{eqa2}) and (\ref{eqb2}). For $\langle \hat{a}^{\dagger}_2 \hat{b}_2 \rangle$ we have
\begin{equation}
\hat{a}^{\dagger}_2 \hat{b}_2 = \hat{a}^{\dagger}_1 e^{-i (\lambda_{bb} - \lambda_{ab}) \hat{b}^{\dagger}_1 \hat{b}_1} e^{i (\lambda_{aa} - \lambda_{ab}) \hat{a}^{\dagger}_1 \hat{a}_1} \hat{b}_1 
\end{equation}
To calculate the expectation value of this expression, the creation and annihilation operators must be normally ordered. Making use of the identity \cite{louisell}
\begin{equation}
e^{x \hat{a}^{\dagger} \hat{a}} = \sum_{n=0}^{\infty} \frac{(e^x - 1)^n} {n!} (\hat{a}^{\dagger})^n (\hat{a})^n
\end{equation}
allows us to compute expectation values of coherent states of the form
\begin{eqnarray}
\langle \alpha | e^{x \hat{a}^{\dagger} \hat{a}} |\alpha \rangle &=& \sum_{n=0}^{\infty} \frac{(e^x - 1)^n} {n!} |\alpha|^{2n} \nonumber \\
%
&=& \exp[|\alpha|^2(e^x -1)] \label{eqExpectationValueOfExponential}
\end{eqnarray}
This gives
\begin{eqnarray}
\langle \hat{a}^{\dagger}_2 \hat{b}_2 \rangle &=& \langle \alpha, \beta | \hat{a}^{\dagger}_1 e^{-i (\lambda_{bb} - \lambda_{ab}) \hat{b}^{\dagger}_1 \hat{b}_1} e^{i (\lambda_{aa} - \lambda_{ab}) \hat{a}^{\dagger}_1 \hat{a}_1} \hat{b}_1 | \alpha, \beta \rangle \nonumber \\
%
&=& \alpha^* \beta \exp[|\alpha|^2(e^{i (\lambda_{aa} - \lambda_{ab})} -1)] \exp[|\beta|^2 (e^{-i (\lambda_{bb} - \lambda_{ab})} -1)]
\end{eqnarray}

@@

\langle \hat{a}^{\dagger}_2 \hat{a}^{\dagger}_2 \hat{b}_2 \hat{b}_2 \rangle = \alpha^{*2} \beta^2 e^{i(\lambda_{aa}-\lambda_{bb})} \exp[|\alpha|^2(e^{2i (\lambda_{aa} - \lambda_{bb})} -1)] \exp[|\beta|^2(e^{-2i (\lambda_{bb} - \lambda_{bb})} -1)] 

@@

&& + i c s^3 [2 n_a e^{i(\lambda_{aa}-\lambda_{ab})}] A^* B - 2 n_a e^{-i(\lambda_{aa}-\lambda_{ab})} A B^* - A^* B + A B^*] 

@@

When substituted into (\ref{eqNbvariance}) and (\ref{eqNbvariance}) these expecation values give the absolute number variances stated in Section \ref{secTwoModeAnalytic}. To calculate the number difference variance, we note

@@

{\mathrm{Var}}[N_a-N_b] = {\mathrm{Var}}[N_a] + {\mathrm{Var}}[N_b] - 2\langle \hat{a}^{\dagger}_2 \hat{a}_2 \hat{b}^{\dagger}_2 \hat{b}_2\rangle + 2\langle \hat{a}^{\dagger}_2  \hat{a}_2 \rangle \langle \hat{b}^{\dagger}_2 \hat{b}_2\rangle \label{eqNumDiffVarianceAppendix}.

@@

Substituting the expectation values calculated earlier in this Appendix into (\ref{eqNumDiffVarianceAppendix}) and simplifying yields the expression (\ref{eqNumDiffVariance}) given in Section \ref{secTwoModeAnalytic}.

>>>>

\appendix
\section{Derivation of two mode squeezing formula}
\label{appendixTwoModeDerivation}
In this Appendix we derive the equations for the number variances which were earlier stated without proof. We use the model described in Section \ref{secTwoModeAnalytic}, where we consider a two mode system with $N$ particles, initially prepared as a coherent state with all the population in mode $|a\rangle$ at time $t=0$. A coupling pulse is then applied, so that the system evolves under the Hamiltonian
\begin{equation}
\hat{H_1} = \hbar \omega \hat{a}^{\dagger} \hat{a} +  \hbar \omega \hat{b}^{\dagger} \hat{b} 
          + \Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger} \hat{a} )
\end{equation}
until time $t=t_1$. From time $t_1$ until time $t=t_2$ the coupling is turned off, and the system evolves under the Hamiltonian
\begin{equation}
\hat{H_2} = \frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}.
\end{equation}
Finally, from time $t_2$ until time $t=t_3$, the coupling is once again applied with a phase offset of $\phi$ relative to the first time, and the system evolves via
\begin{equation}
\hat{H_3} = \hbar \omega \hat{a}^{\dagger} \hat{a} +  \hbar \omega \hat{b}^{\dagger} \hat{b}
          + \Omega (e^{i\phi} \hat{a}^{\dagger} \hat{b} + e^{-i\phi} \hat{b}^{\dagger} \hat{a} ).
\end{equation}
It is clear that the linear self-energy terms proportional to $\omega$ result in a constant phase rotation of the fields $\hat{a}$ and $\hat{b}$. As this system is self-contained and global phase is unmeasurable, all phase measurements can only rely on the difference in phase between the two modes. This phase rotation is therefore irrelevent and will be ignored in the following analysis.

@@

Beginning with the Hamiltonian $H_1$, during the period $(0, t_1)$ mode $\hat{a}$ will evolve as
\begin{equation}
\hat{a}_1 = e^{ i H_1 t/ \hbar} \, \hat{a}_0 \, e^{-i H_1 t/ \hbar}.
\end{equation}
Utilizing the Hadamard lemma
\begin{equation}
e^X Y e^{-X} = Y + [X,Y] + \frac{1}{2!}[X,[X,Y]] + \frac{1}{3!}[X,[X,[X,Y]]] + \ldots
\label{eqHadamard}
\end{equation}
we obtain 
\begin{eqnarray}
\hat{a}_1 &=& \hat{a}_0 - i \Omega t_1 \hat{b}_0 + \frac{(i \Omega t_1)^2}{2!} \hat{a}_0 - \frac{(i \Omega t_1)^3}{3!} \hat{b}_0 + \frac{(i \Omega t_1)^4}{4!} \hat{a}_2 - \ldots \nonumber \\
          &=& \cos (\Omega t_1) \hat{a}_0 -i \sin (\Omega t_1) \hat{b}_0.
\label{eqa1}
\end{eqnarray}
Similarly, we find
\begin{equation}
\hat{b}_1 = \cos (\Omega t_1) \hat{b}_0 - i \sin (\Omega t_1) \hat{a}_0.
\label{eqb1}
\end{equation}
From the form of (\ref{eqa1}) and (\ref{eqb1}) it is clear that as both modes $\hat{a}$ and $\hat{b}$ began in coherent states, they will remain in coherent states, with only their amplitudes changing. Specifically, at time $t_1$ the system is in a product of coherent states, which we denote $|\alpha, \beta\rangle$ with $\alpha=\sqrt{n_a}$ and $\beta=-i\sqrt{n_b}$.

@@

          e^{-i \lambda_{ab} \hat{a}_1^{\dagger} \hat{a}_1 \hat{b}_1^{\dagger} \hat{b}_1}

@@

   &=& \exp[-i \lambda_{aa} \hat{a}_1^{\dagger} \hat{a}_1] \hat{a}_1.

@@

{\mathrm{Var}}[N_b] &=& \langle \hat{b}^{\dagger}_3 \hat{b}_3 \hat{b}^{\dagger}_3 \hat{b}_3 \rangle - \langle \hat{b}^{\dagger}_3 \hat{b}_3 \rangle ^2.

@@

\hat{a}^{\dagger}_3 \hat{a}_3 &=& \hat{a}^{\dagger}_2 \hat{a}_2 c^2 +  \hat{b}^{\dagger}_2 \hat{b}_2 s^2 + i c s (e^{-i \phi} \hat{b}^{\dagger}_2 \hat{a}_2 - e^{i \phi} \hat{a}^{\dagger}_2 \hat{b}_2) \\

@@

\hat{b}^{\dagger}_3 \hat{b}_3 &=& \hat{b}^{\dagger}_2 \hat{b}_2 c^2 +  \hat{a}^{\dagger}_2 \hat{a}_2 s^2 + i c s ( e^{i \phi} \hat{a}^{\dagger}_2 \hat{b}_2 - e^{-i \phi} \hat{b}^{\dagger}_2 \hat{a}_2) \\

@@

As $\hat{a}^{\dagger}_1 \hat{a}_1$ and $\hat{b}^{\dagger}_1 \hat{b}_1$ commute with the Hamiltonian $H_2$, they are constants of motion so we have
\begin{equation}
\langle \hat{a}^{\dagger}_2 \hat{a}_2 \rangle = \langle \hat{a}^{\dagger}_1 \hat{a}_1 \rangle = \langle \alpha, \beta | \hat{a}^{\dagger}_1 \hat{a}_1 |\alpha, \beta \rangle = |\alpha|^2 = n_a
\end{equation}
%
\begin{equation}
\langle \hat{b}^{\dagger}_2 \hat{b}_2 \rangle = \langle \hat{b}^{\dagger}_1 \hat{b}_1 \rangle = \langle \alpha, \beta | \hat{b}^{\dagger}_1 \hat{b}_1 |\alpha, \beta \rangle = |\beta|^2 = n_b
\end{equation}
%
\begin{equation}
\langle \hat{a}^{\dagger}_2 \hat{a}_2 \hat{b}^{\dagger}_2 \hat{b}_2 \rangle = \langle \hat{a}^{\dagger}_1 \hat{a}_1 \hat{b}^{\dagger}_1 \hat{b}_1 \rangle = \langle \alpha, \beta | \hat{a}^{\dagger}_1 \hat{a}_1 \hat{b}^{\dagger}_1 \hat{b}_1 |\alpha, \beta \rangle = |\alpha|^2 |\beta|^2 = n_a n_b
\end{equation}
Furthermore, $\hat{a}^{\dagger}_2 \hat{a}_2 \hat{a}^{\dagger}_2 \hat{a}_2$ and $\hat{b}^{\dagger}_2 \hat{b}_2 \hat{b}^{\dagger}_2 \hat{b}_2$ also commute with $H_2$, giving
\begin{equation}
\langle \hat{a}^{\dagger}_2 \hat{a}_2 \hat{a}^{\dagger}_2 \hat{a}_2 \rangle = \langle \hat{a}^{\dagger}_1 \hat{a}_1 \hat{a}^{\dagger}_1 \hat{a}_1 \rangle = \langle \alpha, \beta | \hat{a}^{\dagger}_1 \hat{a}^{\dagger}_1 \hat{a}_1 \hat{a}_1 + \hat{a}^{\dagger}_1 \hat{a}_1 |\alpha, \beta \rangle = n_a^2 + n_a
\end{equation}
%
\begin{equation}
\langle \hat{b}^{\dagger}_2 \hat{b}_2 \hat{b}^{\dagger}_2 \hat{b}_2 \rangle = \langle \hat{b}^{\dagger}_1 \hat{b}_1 \hat{b}^{\dagger}_1 \hat{b}_1 \rangle = \langle \alpha, \beta | \hat{b}^{\dagger}_1 \hat{b}^{\dagger}_1 \hat{b}_1 \hat{b}_1 + \hat{b}^{\dagger}_1 \hat{b}_1 |\alpha, \beta \rangle = n_b^2 + n_b
\end{equation}
The remaining terms are not constants of motion, so we proceed by making use of (\ref{eqa2}) and (\ref{eqb2}). For $\langle \hat{a}^{\dagger}_2 \hat{b}_2 \rangle$ we have
\begin{equation}
\hat{a}^{\dagger}_2 \hat{b}_2 = \hat{a}^{\dagger}_1 e^{-i (\lambda_{bb} - \lambda_{ab}) \hat{b}^{\dagger}_1 \hat{b}_1} e^{i (\lambda_{aa} - \lambda_{ab}) \hat{a}^{\dagger}_1 \hat{a}_1} \hat{b}_1. 
\end{equation}
To calculate the expectation value of this expression, the creation and annihilation operators must be normally ordered. Making use of the identity \cite{louisell}
\begin{equation}
e^{x \hat{a}^{\dagger} \hat{a}} = \sum_{n=0}^{\infty} \frac{(e^x - 1)^n} {n!} (\hat{a}^{\dagger})^n (\hat{a})^n
\end{equation}
allows us to compute expectation values of coherent states of the form
\begin{eqnarray}
\langle \alpha | e^{x \hat{a}^{\dagger} \hat{a}} |\alpha \rangle &=& \sum_{n=0}^{\infty} \frac{(e^x - 1)^n} {n!} |\alpha|^{2n} \nonumber \\
%
&=& \exp[|\alpha|^2(e^x -1)]. \label{eqExpectationValueOfExponential}
\end{eqnarray}
This gives
\begin{eqnarray}
\langle \hat{a}^{\dagger}_2 \hat{b}_2 \rangle &=& \langle \alpha, \beta | \hat{a}^{\dagger}_1 e^{-i (\lambda_{bb} - \lambda_{ab}) \hat{b}^{\dagger}_1 \hat{b}_1} e^{i (\lambda_{aa} - \lambda_{ab}) \hat{a}^{\dagger}_1 \hat{a}_1} \hat{b}_1 | \alpha, \beta \rangle \nonumber \\
%
&=& \alpha^* \beta \exp[|\alpha|^2(e^{i (\lambda_{aa} - \lambda_{ab})} -1)] \exp[|\beta|^2 (e^{-i (\lambda_{bb} - \lambda_{ab})} -1)].
\end{eqnarray}

@@

\langle \hat{a}^{\dagger}_2 \hat{a}^{\dagger}_2 \hat{b}_2 \hat{b}_2 \rangle = \alpha^{*2} \beta^2 e^{i(\lambda_{aa}-\lambda_{bb})} \exp[|\alpha|^2(e^{2i (\lambda_{aa} - \lambda_{bb})} -1)] \exp[|\beta|^2(e^{-2i (\lambda_{bb} - \lambda_{bb})} -1)].

@@

&& + i c s^3 [2 n_a e^{i(\lambda_{aa}-\lambda_{ab})}] A^* B - 2 n_a e^{-i(\lambda_{aa}-\lambda_{ab})} A B^* - A^* B + A B^*]. 

@@

When substituted into (\ref{eqNavariance}) and (\ref{eqNbvariance}) these expecation values give the absolute number variances stated in Section \ref{secTwoModeAnalytic}. To calculate the number difference variance, we note

@@

{\mathrm{Var}}[N_a-N_b] = {\mathrm{Var}}[N_a] + {\mathrm{Var}}[N_b] - 2\langle \hat{a}^{\dagger}_3 \hat{a}_3 \hat{b}^{\dagger}_3 \hat{b}_3\rangle + 2\langle \hat{a}^{\dagger}_3  \hat{a}_3 \rangle \langle \hat{b}^{\dagger}_3 \hat{b}_3\rangle \label{eqNumDiffVarianceAppendix}.

@@

We write (\ref{eqNumDiffVarianceAppendix}) in terms of $\hat{a}_2$ and $\hat{b}_2$ by using using (\ref{eqa3}) and (\ref{eqb3}), and then use the expectation values calculated earlier in this Appendix. After some algebra we obtain the expression (\ref{eqNumDiffVariance}) given in Section \ref{secTwoModeAnalytic}.
